\documentclass[a4paper,10pt]{amsbook}
\usepackage{fullpage}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{newverbs}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{enumitem}

\input{macros}

% Create a new list environment
\newlist{secenum}{enumerate}{1}

% Configure it to use the custom counter without resetting it at \begin{secenum}
\setlist[secenum,1]{
    label=\thesection.\arabic*,
    ref=\thesection.\arabic*
}


\begin{document}
\begin{center}
 {\Large Maths with Python --- Python reference}
\end{center}
\vspace{4ex}

\section{Getting started}
\label{sec-start}

These instructions will assume that you are comfortable with using
Python in Jupyter notebooks in VS Code, and that you have installed 
the packages \pcode+sympy+, \pcode+numpy+, \pcode+scipy+ and
\pcode+plotly+ using \pcode+pip+ or \pcode+conda+.  You will also 
need to import these packages and set some options.  The most
convenient way to do this is as follows.  Distributed together with
this file is a directory called \pcode+header+ containing a single
file \pcode+__init__.py+.  You should copy this directory to be a
subdirectory of the directory where you keep your notebooks for this
course, then you should include the line 
\begin{pcodeblock}
 from header import *
\end{pcodeblock}
as the top line of each of your notebooks.

\section{Notation for variables}

\begin{secenum}
 \item\label{symbols} By default names like $x$ refer to variables as used
  in programming languages like Python.  They are expected to have a
  value (which might be a number or a list or something more
  complicated).  If you try to use a variable (for example, by trying
  to expand $(1+x)^4$) without assigning a value  then you will get an
  error.  If you instead want to treat $x$ as an abstract symbol 
  then you need to enter \pcode+x = sp.symbols('x')+.  You can do this for 
  several variables at the same time using syntax like 
  \pcode+x,y,a,b = sp.symbols('x y a b')+.  
%
 \item\label{subscripts} If you want variables 
  with indices like $m_0,m_1,m_2$ then you should enter 
  \pcode+m = sp.IndexedBase('m')+, and then enter \pcode+m[2]+ for
  $m_2$, for example.
%
 \item\label{variables} Most variables will have single-letter
  names, such as $x$ or $A$.   Note that case is significant:
  Python regards $A$ as different from $a$, and $x$ as different
  from $X$.  It is perfectly permissible to use long
  names for variables, such as \pcode~eqns~ or \pcode~sols~ or
  \pcode~SpeedOfLight~.  However, you may not use words that
  already have a special meaning in Python.  In particular,
  you cannot call a variable \pcode~lambda~. 
%
  \item\label{greek} If you enter \pcode+mu = sp.symbols('mu')+ then in some
  (but not all) contexts, \pcode+mu+ will be displayed as $\mu$.
   Alternatively, there are various ways to set things up so that you
   can enter Greek letters directly (one option is 
   \href{Wincompose}{https://wincompose.info/}).  Then you can just 
   use Greek letters like normal English letters.
\end{secenum}

\section{Generalities}

\begin{secenum}
\item\label{help} The \pcode+help()+ command will give you details about any
 Python or SymPy object that you are interested in.  Alternatively, in
 VS Code you can just enter a name and hover over it with your mouse
 and some information will appear.  However, the output from \pcode+help()+
 is sometimes very long and overwhelming.  It may be better to ask an
 AI a more specific question about what you are hoping to do.
%
 \item\label{problems} See Section~\ref{sec-problems}
 for advice on some common problems. 
%
 \item\label{lastoutput} If you have several commands in the same code cell
 then Python will only print the result of the last one by default.
 If you want to see other things then you should add explicit
 \pcode+print()+ or \pcode+display()+ commands.
%
 \item\label{semicolon}
 If you end a command with a semicolon then Python will not 
 print the result even if it is the last command in the cell.
 You can also achieve the same effect by entering \pcode+None+
 as the last command in the cell.
%
 \item\label{underscore} You can use the symbol \pcode~_~
 to refer to the last thing that Python calculated and displayed,
 and \pcode~__~ to refer to the one before that, and so on. 
 However, this mechanism cannot be used to refer to things that were
 calculated but not displayed.
 If you move around a worksheet inserting things in
 different places, you should remember that \pcode~_~ refers
 to the \textbf{most recent} result, which may not be the
 one that you see directly above the line where you are
 typing.  
%
 \item\label{equals}
 It is important to understand the difference between
 \pcode~=~ and \pcode~==~ and \pcode+sp.Eq()+.  
 \begin{itemize}
  \item If we enter \pcode~x=3~, this defines $x$ to be equal to $3$, so
   Python will replace $x$ by $3$ whenever it sees an $x$.
  \item If we instead enter \pcode~x==3~, we are checking whether $x$ is
   currently straightforwardly equal to $3$.  If $x$ is an abstract
   symbol which has not been assigned a numerical value then \pcode~x==3~
   will just be \pcode~False~, even if $x$ could be equal to $3$ in some
   contexts.  
  \item Assuming that \pcode~a~ and \pcode~b~ have been declared to be
   abstract symbols, the expression\\ \pcode~sp.Eq(2 * a, b + 1)~
   represents the hypothetical question ``Is $2 a$ equal to $b + 1$?'',
   which might be true or false depending on the values of $a$ and
   $b$.  Entering this question does not assign any values to
   \pcode~a~ or \pcode~b~, and does not return a \pcode~True~ or
   \pcode~False~ answer.  It simply represents the question in a
   symbolic form.  By contrast, the expression \pcode~2 * a == b + 1~
   gives \pcode~False~, because the expression \pcode~2 * a~ is not
   visibly the same as \pcode~b + 1~.  In fact, even
   \pcode~(a + 1) * (a - 1) == a ** 2 - 1~ will return \pcode~False~,
   because the two sides of the equation are not visibly the same,
   even though they become the same after expanding out the left hand
   side.  (The right way to test whether $A$ and $B$ are
   mathematically equivalent is to enter \pcode~sp.simplify(A-B)~ and
   see whether you get zero.)  The main use for \pcode~sp.Eq()~ is to
   introduce equations which you later intend to solve.
 \end{itemize}
%
 \item\label{funcdef}
 If you want to define a function (say $f(x)=x^2+3$) it does
 \emph{not} work to enter \pcode~f(x)=x^2+3;~.  See
 Section~\ref{sec-funcdef}. 
%
 \item\label{restart}
  Near the top of your VS Code window you should see a menu item
  marked \pcode+Restart+.  If you click this then Python will restart
  and all values that you might have assigned to variables will be
  removed.  You might want to do this at the beginning of each new
  exercise, otherwise stray values left over from previous exercises
  can cause trouble and confusion.  Note, however, that most notebooks
  will have a cell at the top with some \pcode+import+ statements.
  Restarting removes all imports, so you will need to go back to your
  import cell and execute it again before continuing.
%
 \item\label{unassign}
  If you just want to clear a single variable (say $x$) you
  can enter \pcode~del x~ instead.  To clear several variables, you
  can enter something like \pcode+del A,x,U+.
\end{secenum}

\section{Numerical approximation}

\begin{secenum}
 \item\label{float} To get an explicit numerical approximation for an
  expression, use the function \pcode~sp.N()~ function.  For example,
  \pcode~sp.N(sp.pi)~ gives $3.14159265358979$, and
  \pcode~sp.N(x/3)~ gives $0.3333333333 x$. 
%
 \item\label{floatdigits} The function \pcode~sp.N()~ generally gives an
  answer to $15$ significant figures.  If you want to
  calculate $\pi^2/6$ to $50$ digits (for example), you can
  enter \pcode~sp.N(sp.pi**2/6, 50)~.
%
 \item\label{digitslost} It is possible to lose accuracy for non-obvious
 reasons.  For example, if you enter \pcode+sp.N(1/7, 100)+ then
 Python will work from the inside out, first calculating $1/7$ to low
 accuracy and only then calling \pcode+sp.N()+ with a request for high
 accuracy that can no longer be achieved.  One way to fix this is to
 enter \pcode+sp.N(1,100)/7+, which requests high accuracy before
 attempting to divide by $7$.
\item\label{exactrational} In some contexts, SymPy will convert
 fractions to approximate decimals without being asked, which is not
 what we want when doing symbolic mathematics.  To prevent $2/3$ from
 being converted to $0.66666666667$, we need to enter it as
 \pcode~sp.Rational(2, 3)~ instead of \pcode~2/3~.  Sometimes this is not
 necessary, but the rules for whether it is necessary are complicated,
 so it is best to err on the safe side.
\end{secenum}


\section{Algebraic operations}

\begin{secenum}
 \item\label{star} Use a \pcode~*~ for multiplication; for example,
 $2xy$ should be entered as \pcode~2*x*y~, and $(2x+1)\sin(3x)$
 should be entered as \pcode~(2*x+1)*sp.sin(3*x)~. 
%
 \item\label{twoletters}
 Note also that Python will treat \pcode~xy~ as a two-letter
 name for a single variable; if you want $x$ times $y$, you
 must enter \pcode~x*y~. 
%
\item\label{timesbrackets} Brackets need to be used for algebraic
 grouping, just as in ordinary mathematical writing.  For the product
 of $x + 2$ and $x - 2$ we need to enter \pcode~(x + 2) * (x - 3)~,
 not \pcode~x + 2 * x - 2~.  Brackets used for algebraic grouping
 should always be round brackets, not square or curly ones.
%
 \item\label{powers} Enter powers using \pcode~**~,
 for example \pcode~x ** 6~ for $x^6$.  The notation \pcode+x^6+ does
 have a meaning (the exclusive or operator applied to $x$ and $6$) but
 it is almost never what you want; this can unfortunately lead to some
 confusing error messages.
%
\item\label{powerbrackets} Bracketing is also important when raising
 things to powers.  To raise $x + 2$ to the power $n+5$, we need to
 enter \pcode~(x + 2) ** (n + 5)~, not \pcode~x + 2 ** n + 5~.  The
 cube root of $x$ is $x^{1/3}$.  For this to work properly in SymPy,
 it is important to enter $1/3$ as \pcode~sp.Rational(1,3)~ (which
 gives $1/3$ exactly) and not $1/3$ (which gives 0.33333).
%
\item\label{sqrt} The (nonnegative) square root of $x$ can be entered
 as \pcode~sp.sqrt(x)~ or \pcode~x ** sp.Rational(1,2)~.  If you enter
 \pcode~x ** (1/2)~ you will only get an approximate answer.
\end{secenum}

\section{Algebraic manipulation}

Suppose we have a complicated expression called $A$, which we want to
simplify or otherwise manipulate. 
\begin{secenum}
 \item\label{subs} To substitute for some or all of the variables in
  $A$, use the \pcode~subs~ command.  For example, if $A$ is
  $(a+b)^n$, you can change the $b$ to $1$ and the $n$ to
  $p/q$ by entering \pcode~A.subs({b:1,n:p/q})~, which gives
  $(a+1)^{p/q}$.  (Note that the correct syntax here is \pcode~b:1~,
  not \pcode~b=1~.)
%
 \item\label{solvesubs}
  Often the equations to be used in \pcode~subs(...)~ will
  come from the \pcode~solve(...)~ command.  For example, to
  find the value of $x^2+y^2$ at the point where $x+y=6$ and
  $x-y=2$, enter \pcode~sol=sp.solve([sp.Eq(x+y,6),sp.Eq(x-y,2)])~ 
  and then \pcode~(x**2 + y**2).subs(sol)~.
%
 \item\label{expand} Enter \pcode~sp.expand(A)~ to expand everything out.  For
  example, enter \pcode~sp.expand((x+y)**3)~ to expand
  $(x+y)^3$, giving $x^3+3x^2y+3xy^2+y^3$. 
%
 \item\label{simplify} To simplify $A$, try entering
  \pcode~sp.simplify(A)~.  If this 
  does not do what you want, you can try \pcode~sp.factor(A)~. 
  If $A$ involves trigonometric functions, try
  \pcode~sp.trigsimp(A)~ or \pcode~sp.expand_trig(A)~. 
%
 \item\label{symbolic} Note that rules like $(a^4)^{1/4}=a$ are
  \emph{not valid} when $a$ is negative.  The function
  \pcode~sp.simplify()~ will not simplify expressions like this by
  default.  However, if you declare $a$ to be positive using
  \pcode~a = sp.symbols('a', positive=True)~ (instead of just
  \pcode~a = sp.symbols('a')~) then SymPy will apply rules that are
  only valid for positive $a$.  For example,
  \pcode~(a ** 4) ** sp.Rational(1,4)~ will simplify to $a$.  There
  are also more complex ways to specify that $a$ is an integer or lies
  in a certain range or has other useful properties that could be used
  when simplifying; see the SymPy documentation about assumptions.
%
 \item\label{testequal} If you suspect that $A$ is equal to some other expression (say
  $B$) and want to check this, the best way is to enter
  \pcode~sp.simplify(A-B)~ and see if you get zero. 
%
 \item\label{collect} To collect together all the terms in $A$ involving the same
  power of $x$, enter \pcode~A.collect(x)~.  For example, this converts
  $1+ax+bx+cx^2+dx^2$ to $1+(a+b)x+(c+d)x^2$. 
%
 \item\label{coeff}
  If you just want the
  coefficient of $x^2$ in $A$ then you can instead enter \pcode~A.coeff(x,2)~
  or \pcode~A.coeff(x**2)~.  To get the constant term, you
  must enter \pcode~A.coeff(x,0)~ --- the syntax
  \pcode~A.coeff(x**0)~ will not work. 
\end{secenum}

\section{Constants}

\begin{secenum}
 \item\label{rational} If you enter a fraction like \pcode~2/9~ then it
  will immediately be converted to an approximate decimal, which is 
  not usually what you want.  If you want to treat it as an exact 
  fraction you should enter \pcode~sp.Rational(2,9)~ instead.
%
 \item\label{pi} The constant $\pi\simeq 3.1415926536$ should be entered as
  \pcode~sp.pi~.  If you just enter \pcode~pi~ then that will just 
  be treated as the name of a variable with no special value or
  properties.  Thus, if you have not introduced \pcode~pi~ in any way
  then you will get an error. 
%
 \item\label{exp} The constant $e\simeq 2.718281828$ must be entered as 
  \pcode~sp.E~ (with a capital E) or \pcode~sp.exp(1)~.  
%
 \item\label{imaginary} The square root of minus one can be
  entered as \pcode~sp.I~.
%
 \item\label{infinity} In many contexts $\infty$ can be entered as
 \pcode~sp.oo~.  For example, we can enter \\
 \pcode~sp.limit((1 + 2*x)/(3 + 4*x), x, sp.oo)~ to find the 
 limit of $(1+2x)/(3+4x)$ as $x\to\infty$.
\end{secenum}

\section{Solving}

\begin{secenum}
%
 \item\label{solve} To solve the equation $x^2+6=5x$ (for example), enter
  \pcode~sp.solve(sp.Eq(x**2+6,5*x))~.  The answer is $[2,3]$,
  indicating that the solutions are $x=2$ and $x=3$.   
  If you want to solve for an
  expression being equal to zero then you do not need \pcode~sp.Eq()~,
  you can just enter \pcode~sp.solve(x**2+6-5*x)~ for example.  
  Similarly, if you have already defined a function $g$, you can enter
  \pcode~solve(g(t),t)~ to solve the equation $g(t)=0$. It
  \textbf{does not work} to enter \pcode~a == b~ instead of
  \pcode~sp.Eq(a, b)~.  
%
  \item\label{roots}
   The syntax discussed above gives the solution to $x^2+6-5x=0$ in the form
   \pcode+[2,3]+.  If you prefer to have the answer in
   the form \pcode+[{x:2},{x:3}]+ (which can be more convenient if you
   want to substitute the result into some other expression), you
   should instead enter \pcode~sp.solve({x**2-5*x+6},dict=True)~.
%
  \item\label{realroots}
   The approach discussed above will give complex roots as well as real roots.  
   For example, if we declare \pcode+x = sp.symbols('x')+ and then enter 
   \pcode~sp.solve(x**3+x,x)~ we get $[0,-i,i]$.  If we only want real 
   roots we can either use \pcode~sp.real_roots(x**3+x,x)~ or we can
   declare $x$ using \pcode+x = sp.symbols('x',real=True)+ and then
   just use \pcode+sp.solve()+; this will give $x=0$ as the only real root.
%
 \item\label{multisolve} To solve several equations in several variables, put the
  equations in one set of square brackets, and the variables in another
  set.  For example, to solve $2x+3y=5$ and $3x+4y=7$ for $x$ and $y$,
  enter 
\begin{pcodeblock}
 sp.solve([2*x+3*y-5,3*x+4*y-7],[x,y])
\end{pcodeblock}
%
 \item\label{nosols} If there are no solutions, \pcode~sp.solve()~ will
 return the empty list \pcode~[]~.    For example, this happens if we
 try to solve the simultaneous equations $x+y=1$ and $2x+2y=10$. 
%
 \item\label{allsolutions} If the problem involves trigonometric
  functions, then there will typically be infinitely many
  solutions, but \pcode~sp.solve()~ will only list a few of them.  
  To get all solutions, you can try using \pcode~sp.solveset()~
  instead.  For example, we can solve the equation $\cos(x)=1$ by entering
  \pcode+sp.solveset(sp.cos(x)-1,x)+.  The result will be printed as
  $\{2n\pi\;\mid\;n\in\mathbb{Z}\}$, indicating that $\cos(x)=1$ if
  and only if $x$ is an integer multiple of $2\pi$.  There is quite a
  long story about how this result is represented internally in
  Python, which you can read at 
  \url{https://docs.sympy.org/latest/modules/sets.html}.
%
 \item\label{fsolve} To get approximate numerical solutions, you can 
  use \pcode~sp.nsolve()~ instead of \pcode~sp.solve()~.  For example,
  you can enter \pcode~sp.nsolve(sp.cos(x)-sp.cos(x+1),x,42)~ to find
  a number $x$ with $\cos(x)=\cos(x+1)$, starting the search at
  $x=42$.  When using \pcode~sp.nsolve()~, you always need to specify
  the starting point.  If you need to solve a large number of
  equations numerically then it will usually be best to use the
  \pcode~scipy~ or \pcode~numpy~ packages; you should only use
  \pcode~sp.nsolve()~ for occasional numerical solutions when you are
  mostly working with symbolic equations.
% 
 \item\label{identity} Sometimes we need to solve problems like
  this: find $a$ and $b$ such that $a(x-1)^2+b(x+1)^2=x$
  \emph{for all x}.  For this we can enter 
  \begin{pcodeblock}
   sp.solve_undetermined_coeffs(a*(x-1)**2+b*(x+1)**2-x,[a,b],x)
  \end{pcodeblock}
  Similarly, to find $u$ such that $\sin(t+u)=\cos(t)$ for all
  $t$, we could try 
  \begin{pcodeblock}
   sp.solve_undetermined_coeffs(sp.Eq(sp.sin(t+u),sp.cos(t)),[u],t)
  \end{pcodeblock}
  This does not actually work, but we can make it work by giving the
  solver a hint as follows:
  \begin{pcodeblock}
   sp.solve_undetermined_coeffs(sp.Eq(sp.expand_trig(sp.sin(t+u)),sp.cos(t)),[u],t)
  \end{pcodeblock}
\end{secenum}

\section{Standard functions}

\begin{secenum}
%
 \item\label{standard} Most functions can be entered using their
  usual names prefixed with \pcode~sp.~, for example
 \pcode~sp.sin(x)~ or \pcode~sp.ln(y)~.  If you do not want to type
 the prefix, you can change your initial import statement to include 
 \begin{pcodeblock}
  from sympy import sin, cos, tan, exp, ln
 \end{pcodeblock}
 (for example) then you can just enter \pcode~sin(x)~ instead of
 \pcode~sp.sin(x)~. 
%
 \item\label{funcbrackets}
  However, you must always use brackets: enter
  \pcode~sp.tan(x)~ instead of \pcode~tan x~, and \pcode~sp.sin(2*x)~
  instead of \pcode~sin 2x~. 
%
 \item\label{abs} The absolute value of $x$, traditionally written as
  $|x|$, must be entered as \pcode~abs(x)~.  (This is an exception to
  the usual pattern: \pcode+sp.abs(x)+ will not work.)
%
 \item\label{log} Both \pcode~sp.ln(x)~ and \pcode~sp.log(x)~ refer to the natural
  logarithm (to base $e$) of $x$. 
%
 \item\label{logbase}
  If you want to use the logarithm to base $10$ then you
  should enter \pcode~sp.log(x, 10)~.  Similarly, enter
  \pcode~sp.log(x, 2)~ for $\log_2(x)$ (the logarithm to base $2$) and so on. 
%
 \item\label{expx} The function $e^x$ should be entered as \pcode~sp.exp(x)~. 
%
 \item\label{hyp} The hyperbolic function $\sinh(x)=(e^x-e^{-x})/2$ can be
  entered as \pcode~sp.sinh(x)~, and similarly for the functions
  $\cosh(x)=(e^x+e^{-x})/2$ and $\tanh(x)=\sinh(x)/\cosh(x)$.
  If we have an expression $A$ which involves hyperbolic functions, we
  can rewrite it in terms of exponentials using
  \pcode~A.rewrite(sp.exp)~. 
%
 \item\label{sinsquared} The traditional notation $\sin^2(x)$
  refers to the square of $\sin(x)$, which could also be
  written $(\sin(x))^2$ or $\sin(x)^2$.  In Python this must 
  be entered as \pcode~sp.sin(x)**2~.  Similar comments apply to
  $\tan^2(x)$ and so on.  Note that $\sin(x^2)$ means something
  different again: it is the sin of the square of $x$, not the square
  of the sin. 
%
 \item\label{cosec} In traditional notation $1/\sin(x)$ is sometimes written as
  $\text{csc}(x)$ or $\text{cosec}(x)$.  In Python \pcode~sp.csc(x)~ will
  work, but \pcode~sp.cosec(x)~ will not.  Of course, you can also just
  enter \pcode~1/sp.sin(x)~.  Similarly $1/\cos(x)$ is \pcode~sp.sec(x)~.
%
 \item\label{arctan} In traditional notation $\tan^{-1}(x)$ or $\arctan(x)$
  or $\text{atan}(x)$ refers to the angle $\theta$ such that
  $\tan(\theta)=x$.  This is completely different from the number
  $\tan(x)^{-1}=1/\tan(x)$.  In Python you should enter
  \pcode~sp.atan(x)~ (\textbf{not} \pcode~tan^(-1)(x)~ or anything
  like that).  Similar comments apply 
  to $\text{asin}(x)$, $\text{atanh}(x)$ and so on. 
%
 \item\label{possqrt} When $x\geq 0$, the notation \pcode~sp.sqrt(x)~ means, by
  definition, the positive square root of $x$.  (There is more to say
  if $x$ is negative or complex, but we pass over that here.)  Thus,
  for example, \pcode~sp.sqrt(9)~ is definitely $3$ and not $-3$.  
\end{secenum}

\section{Defining your own functions}\label{sec-funcdef}

\begin{secenum}
%
 \item\label{arrow} If you want to define a simple function like 
  $f(x)=x^2$, enter \pcode~f = lambda x: x^2~.  It
  \textbf{does not} work to enter \pcode~f(x)=x^2;~ instead.
%
 \item\label{seriesdef}
  Similarly, to define the sequence $a(n)=(-1)^n/(x+n)$, enter
  \pcode~a = lambda n: (-1)**n/(x+n)~, not \\
  \pcode~a(n)=(-1)**n/(x+n)~. 
%
 \item\label{badderiv} It \textbf{does not} work to use notation like
   \pcode~f'(x)~ for the derivative of $f(x)$; see
   Section~\ref{sec-diff} instead. 
%
 \item\label{multifunc}
  You can define a function of several variables in a similar
  way.  For example, to define $g(a,u,v)=u^a+v^a$, enter
  \pcode~g = lambda a,u,v: u**a+v**a~. 
%
 \item\label{longfunc}
  For longer and more complicated functions you may prefer to use
  the more standard Python function definition syntax, like
  \begin{pcodeblock}
   def f(x, y, z):
       return (x+y)**8 + (x-y)**8 + (x+z)**8 + (x-z)**8
  \end{pcodeblock}
\end{secenum}

\section{Plotting}

In this course we mostly use the \pcode+sympy+ library, but that does
not, by itself, support plotting.  We instead use the \pcode+numpy+
and \pcode+plotly+ libraries to make plots.  (There is also another
popular library called \pcode+matplotlib+ which could be used instead
of \pcode+plotly+, but on balance we have found \pcode+plotly+ to be
preferable.)  More details of everything discussed here can be found
at \url{https://plotly.com/python/}.

\begin{secenum}
%
 \item\label{basicplot} To plot the graph $y=x^3-x$ from $x=-2$ to $x=2$ (for
  example), we would enter
  \begin{pcodeblock}
   x = np.linspace(-2, 2, 201)
   y = x**3 - x
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=x, y=y))
   fig.show()
  \end{pcodeblock}
  The first line sets $x$ to be an array
  $[x_0,x_1,x_2,\dotsc,x_{200}]=[-2,-1.98,-1.96,\dotsc,2]$ consisting
  of $201$ equally spaced points between $-2$ and $2$ (inclusive).
  The second line sets $y$ to be the array
  $[y_0,\dotsc,y_{200}]=[x_0^3-x_0,\dotsc,x_{200}^3-x_{200}]$ of
  corresponding $y$ values.  The
  third line \pcode~fig = go.Figure()~ creates an object called
  \pcode+fig+ representing a blank 
  picture, to which we can add curves, shapes, text and so on.  In this
  case we just add a single curve passing through the points with
  coordinates $(x_i,y_i)$; this is done by the line
  \pcode+fig.add_trace(go.Scatter(x=x, y=y))+.  Finally, the line
  \pcode+fig.show()+ shows the picture that we have constructed.
  This is not always necessary, if the last line in a cell is related
  to plotting then Python will often decide for itself that the
  resulting plot should be shown, but it does not hurt to make it
  explicit.  The graph has various interactive features which you can
  discover by hovering and clicking with your mouse. 
% 
 \item\label{legend}
Here is a version with some additional features.  We have added
options to make the curve into a thick orange dotted line.  The option
\pcode~name=r'$y = x^3-x$'~ specifies the name of the curve, but this is not
displayed by default.  The option \pcode~showlegend=True~ specifies that
there should be a box showing the names of all named elements in the
plot.  In this case we have a single named curve, and the legend is
shown at the top right. 

Note that the name is written in LaTeX, and we have written
\pcode~name=r'...'~ instead of just \pcode~name='...'~.  The \pcode~r~
prefix prevents special treatment of the backslash character, and is
generally needed when entering strings containing LaTeX.  In this case
we have no backslashes so the \pcode~r~ is not required, but it is safest to
use it consistently. 

Note also that the colour must be referred to using the US spelling
(\pcode~line_color~ rather than \pcode~line_colour~).  We have specified the
colour as \pcode~orange~, and many other normal colour names will also work.
However, for more precise control one can use hex codes, such as
\pcode~#40e0d0~ for a certain shade of turquoise.  Sites such as
\href{htmlcolorcodes.com}{https://htmlcolorcodes.com} will help you to find
appropriate hex codes. 
\begin{pcodeblock}
    x = np.linspace(-2, 2, 201)
    y = x**3 - x
    fig = go.Figure()
    fig.add_trace(go.Scatter(x=x, y=y, name=r'$y = x^3-x$',
                  showlegend=True, line_color='orange',
                  line_dash='dot', line_width=5))
    fig.show()
\end{pcodeblock}
%
 \item\label{layout} Various features of the overall picture can be changed 
  using \pcode+fig.update_layout()+ (before \pcode+fig.show()+).  
  For example, the code below specifies that the plot should be
  $800\times 400$ pixels, with the $x$ axis running from $x=-3$ to
  $x=3$ and the $y$-axis from $y=0$ to $y=10$. 
  \begin{pcodeblock}
    x = np.linspace(-2, 2, 201)
    y = x**3 - x
    fig = go.Figure()
    fig.add_trace(go.Scatter(x=x, y=y, mode='lines', name=r'$y = x^3-x$', showlegend=True))
    fig.update_layout(height=400, width=800, xaxis_range=[-3,3], yaxis_range=[0,10])
    fig.show()
  \end{pcodeblock}
%
 \item\label{scaling} To make the vertical scale the same as the horizontal
  scale, add the options\\
  \pcode~yaxis_scaleanchor="x",yaxis_scaleratio=1~ 
  in \pcode+fig.update_layout()+.  Alternatively, provided you have done
  \pcode~from header import *~ as suggested at the top of this file,
  you can enter \pcode~equal_aspect(fig)~ to make the scales equal. 
%
 \item\label{twoplots} To plot more than one graph in the same picture, we
  can just call \pcode+fig.add_trace()+ several times.  For example, we
  can plot $y=x^3-x$ and $y=x^3+x$ in the same picture as follows:
  \begin{pcodeblock}
   x = np.linspace(-2, 2, 201)
   y0 = x**3 - x
   y1 = x**3 + x
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=x, y=y0))
   fig.add_trace(go.Scatter(x=x, y=y1))
   fig.show()
  \end{pcodeblock}
  If we do not specify the colours, they will be chosen automatically
  so that the two curves have different colours.  
%
 \item\label{piticks} Similarly, we can plot $\cos(t)$ and $\sin(t)$
  together from $t=-3\pi$ to $t=3\pi$ as shown below.  We have 
  called \pcode~use_pi_ticks(fig,-3,3,2)~ to put tick marks on the
  horizontal axis at positions from $-3\pi$ to $3\pi$ with steps of
  $\pi/2$.  If we did not do that, we would instead get tick marks at
  positions $-8,-6,-4,\dotsc,8$, which is not very helpful when
  plotting functions that are periodic with period $2\pi$.

  The function \pcode~use_pi_ticks()~ will only work if you have done
  \pcode~from header import *~ as suggested at the top of this file.

  Note also that we have used he function \pcode~np.sin()~ (from the
  \pcode~numpy~ library) rather than \pcode~sp.sin()~.  Because
  \pcode~t~ is an array rather than a single number, it would not work
  to use \pcode~sp.sin(t)~.  One could instead use
  \pcode~[sp.sin(t0) for t0 in t]~, but that would not be efficient:
  the \pcode~numpy~ library is highly optimised for numerical
  calculation, and \pcode~sympy~ is not.
 \item\label{hideaxes} To make the $x$ and $y$ axes invisible (together with
  the associated tick marks and labels) add the options
  \pcode+xaxis_visible=False, yaxis_visible=False+ in
  \pcode+fig.update_layout()+. 
 \item\label{discont} If we plot a function with discontinuities, like
  $y=1/x$, then by default we will get spurious vertical lines at the
  discontinuity points.  This can be avoided as follows:
  \begin{pcodeblock}
   x = np.linspace(-2,2,400)
   y = 1/x
   jumps = np.where(np.abs(np.diff(y)) >= 0.5)[0]
   x = np.insert(x, jumps+1, np.nan)
   y = np.insert(y, jumps+1, np.nan)
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=x, y=y, mode='lines', 
                            name=r'$y = \{x\}$', showlegend=True))
   fig.update_layout(height=400, width=800, yaxis_range=[-8,8], showlegend=True)   
  \end{pcodeblock}
  The third line finds the values of $x$ where $y$ jumps by more than
  $0.5$, indicating a discontinuity.  (The value $0.5$ might need to
  be adjusted to give the right outcome in other examples.)  The next
  two lines insert the value \pcode~np.nan~ (indicating ``not a
  number'') at the corresponding places in the arrays $x$ and $y$.
  The plotting system takes this as an indication that the curve
  should be broken without spurious vertical lines.
 \item\label{numpoints} To improve the accuracy of a graph, just increase
  the third argument in \pcode+np.linspace+.  For example, the code
  \begin{pcodeblock}
   x = np.linspace(-1, 1, 50)
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=x, y=x * np.sin(np.pi / x)))
  \end{pcodeblock}
  gives a very jagged picture, but it becomes much better if we change 
  \pcode+np.linspace(-1,1,50)+ to \pcode+np.linspace(-1,1,1000)+.
%
 \item\label{parametric} To plot the curve given parametrically by
  $x=1/(1+t^2)$ and $y=\sin(t)$ from $t=-5$ to $t=5$ (for example), enter
  \begin{pcodeblock}
   t = np.linspace(-5, 5, 400)
   x = 1/(1 + t ** 2)
   y = np.sin(t)
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=x, y=y))
  \end{pcodeblock}
  Similarly, to plot the curve $(x,y)=(\sin(2t),\cos(3t))$
  for $t=0$ to $2\pi$, enter
  \begin{pcodeblock}
   t = np.linspace(0, 2*np.pi, 400)
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=np.sin(2*t), y=np.cos(3*t)))
  \end{pcodeblock}
 \item\label{implicitplot} Consider a curve given implicitly, say
  by the equation $x+y+x^2y^2=1$ with $-3\leq x\leq 3$ and
  $-4\leq y\leq 4$.  This can be plotted as follows:
  \begin{pcodeblock}
   x0 = np.linspace(-3,3,100)
   y0 = np.linspace(-4,4,100)
   x, y = np.meshgrid(x0, y0)
   z = x + y + x**2 * y**2
   fig = go.Figure()
   fig.add_trace(go.Contour(x=x0, y=y0, z=z, 
                            contours=dict(type='constraint', value=1), 
                            line_color='red'))
   fig.update_layout(height=600, width=600, xaxis_range=[-3,3], yaxis_range=[-4,4])
  \end{pcodeblock}
  The first line sets \pcode+x0+ to ba a one-dimensional array of
  $x$-values, and similarly for the second line.  The third line sets
  $x$ to be a two-dimensional array of $x$-values, and $y$ to be a
  two-dimensional array of $y$-values.  The fourth line uses these to
  compute a two-dimensional array $z$ of values of the function
  $x+y+x^2y^2$, then the sixth line draws the curve where $z=1$.
  Similarly, you can plot the circle $x^2+y^2=4$ like this:
  \begin{pcodeblock}
   x0 = np.linspace(-2,2,100)
   y0 = np.linspace(-2,2,100)
   x, y = np.meshgrid(x0, y0)
   z = x**2 + y**2 - 1
   fig = go.Figure()
   fig.add_trace(go.Contour(x=x0, y=y0, z=z, 
                            contours=dict(type='constraint', value=0), 
                            line_color='red'))
   fig.update_layout(height=600, width=600, xaxis_range=[-2,2], yaxis_range=[-2,2])
  \end{pcodeblock}
%
 \item\label{saveplot} 
  Here is an example where we create two separate pictures called
  \pcode~fig1~ and \pcode~fig2~ (which are not displayed), and then combine them
  to make a picture called \pcode~fig~, which is displayed. 
  \begin{pcodeblock}
    x = np.linspace(-1,1,100)
    y = x * (x ** 2 - 1)
    z = 1 - x**4
    fig1 = go.Figure()
    fig1.add_trace(go.Scatter(x=x, y=y, mode='markers', 
                              name=r'$y = x(x^2-1)$', 
                              showlegend=True))
    fig2 = go.Figure()
    fig2.add_trace(go.Scatter(x=x, y=z, mode='lines', 
                              name=r'$y = 1 - x^4$', 
                              showlegend=True))
    fig = go.Figure(data = fig1.data + fig2.data)
    fig.show()
  \end{pcodeblock}
%
 \item\label{listplot}
  By default, \pcode+go.Scatter(x=u, y=v)+ will produce a connected
  curve passing through the points $(u_i,v_i)$.  Sometimes we might
  instead just want to show the points $(u_i,v_i)$ without connecting
  points.  For this, we need to add the option \pcode+mode='markers'+
  to \pcode+go.Scatter()+.  For example, we can display the points 
  $(11,50),(22,40),(33,30),(44,20),(55,10)$ as follows:
  \begin{pcodeblock}
   fig = go.Figure()
   fig.add_trace(go.Scatter(x=[11,22,33,44,55],y=[50,40,30,20,10], mode='markers'))
  \end{pcodeblock}
  Often we just want to display the $y$-values, and the $x$-values are
  not important, we just take them to be $0,1,\dotsc,n-1$.  In that
  context, we do not actually need to specify the $x$-values, we can
  just call something like 
  \pcode+fig.add_trace(go.Scatter(y=[50,40,30,20,10], mode='markers'))+
  and the $x$-values will be inserted automatically.
\end{secenum}

\section{Differentiation}\label{sec-diff}


\begin{secenum}
%
 \item\label{diff} To differentiate $x+\sin(x)$ with respect to $x$ (for
  example), enter \pcode~diff(x+sin(x),x)~.  This is
  equivalent to $\frac{d}{dx}(x+\sin(x))$ in traditional
  notation.  Similarly, to differentiate $1/(1-t^2)$ with
  respect to $t$ we enter \pcode~diff(1/(1-t^2),t)~.  
%
 \item\label{multidiff} To differentiate $y$ three times with respect to $x$
  (for example), enter \pcode~diff(y,x,3)~.  This is equivalent to
  $d^3y/dx^3$ in traditional notation.  Similarly, to find the second
  derivative of $\sin(\theta)$ with respect to $\theta$ we
  enter \pcode~diff(sin(theta),theta,2)~. 
%
 \item\label{implicitdiff} If $x$ and $y$ are related implicitly
  by an equation (say $x+y+x^2y^2=1$) then you can find
  $dy/dx$ in terms of $x$ and $y$ by implicit
  differentiation, like this: 
\begin{pcodeblock}
 sp.idiff(x+y+x^2*y^2-1,y,x)
\end{pcodeblock}
%
\item\label{inertdiff} Sometimes we just want to mention the
  derivative without actually working it out.  We can do this by
  writing \pcode~sp.Derivative()~ instead of \pcode~sp.diff()~.  For example,
  entering \\
  \pcode~u = sp.Derivative(x**2, x)~ gives an object $u$ which is
  displayed as $\frac{d}{dx}x^2$.  If you then enter \pcode~u.doit()~ then
  SymPy will evaluate the derivative and return $2x$.  Below we display
  the unevaluated and evaluated derivatives separately, then we
  illustrate how to make a LaTeX string which displays the equation
  relating them, namely $\frac{d}{dx}x^2 = 2x$.
  \begin{pcodeblock}
    x = sp.symbols('x')
    u = sp.Derivative(x**2,x)
    display(u)
    display(u.doit())
    display(Latex('$' + sp.latex(u) + ' = ' + sp.latex(u.doit()) + '$'))   
  \end{pcodeblock}
  Similarly, \pcode~v = sp.Derivative(t**10,t,2)~ prints as
  $\frac{d^2}{dt^2}t^{10}$, whereas \pcode+v.doit()+ or \\
  \pcode~sp.diff(t**10,t,2)~  gives $90 t^8$. 
%
 \item\label{taylor} To find the Taylor series of a function $f(x)$ near a
  point $x=a$ to order $n$, enter \\
  \pcode~sp.series(f(x),x,a,n)~.  Note that ``order $n$'' means
  that the highest term allowed is a multiple of $x^{n-1}$,
  and all terms $x^n$ and higher are discarded.  For
  example, \pcode~sp.series(1/x,x,2,7)~ gives the Taylor series
  for $1/x$ near $x=2$ including terms up to $(x-2)^6$. 
%
  \item\label{taylorconv} The command \pcode~p = sp.series(...)~ command
  gives an answer including a term like $O(x^7)$, to indicate that
  there are infinitely many more terms, starting with a multiple of
  $x^{7}$.  We cannot do very much with this answer until we have
  converted it to an ordinary polynomial (without the $O(x^7)$ term on
  the end).  We can do this by entering \pcode+p = p.removeO()+.
\end{secenum}

\section{Integration}\label{sec-int}


\begin{secenum}
%
 \item\label{int} To calculate an indefinite integral like
  $\int\tan(x)\,dx$, enter \pcode~sp.integrate(sp.tan(x),x)~. 
%
 \item\label{intconst} Sympy's answer will never include an arbitrary
  constant '$+c$'.  In some contexts (particularly, solution
  of differential equations) the constant makes a
  difference and must be included.  In other contexts it
  is irrelevant; sympy cannot help you to decide about
  this. 
%
 \item\label{defint} To calculate a definite integral like
  $\int_1^5t^3\,dt$, enter \pcode~sp.integrate(t**3,(t,1,5))~. 
  Similarly, to calculate $\int_0^2\sqrt{1+x}\,dx$, enter
  \pcode~sp.integrate(sp.sqrt(1+x),x=0..2)~. 
%
 \item\label{improper} It is also possible to have infinite limits.  For
  example, $\int_{-\infty}^\infty (1+x^4)^{-1}dx$ can be 
  entered as 
  \begin{pcodeblock}
   sp.integrate(1/(1+x**4),(x,-sp.oo,sp.oo))
  \end{pcodeblock}
%
 \item\label{inertint} As with differentiation, it is sometimes useful to
  mention an integral without evaluating it.  You can do
  this by writing \pcode~sp.Integral()~ instead of
  \pcode~sp.integrate()~.  For example, if you enter
  \pcode~u = sp.Integral(x^2,x)~ then you will get an object that is
  displayed as $\int x^2\,dx$; if you then enter
  \pcode~u.doit()~, you will get $x^3/3$. 
\end{secenum}

\section{Sequences and sums}

\begin{secenum}
%
\item\label{seq} We might want to talk about the sequence
 $[1/2, 1/4, 1/6, 1/8, 1/10]$, which is the sequence of terms $1/(2k)$
 for $k$ from $1$ to $5$ inclusive.  There are two ways to do this.
 The first is to use an ordinary Python list comprehension, as shown
 below.  For this we need to use the standard Python convention for
 ranges: the endpoint of the range is never included, so the list
 $1,\dotsc,5$ is \pcode~range(1,6)~ not \pcode~range(1,5)~.

 Although the cell below mentions a variable \pcode~k~, this is actually a
 new temporary variable that only has meaning within the list
 comprehension.  It does not matter whether we already had a variable
 called \pcode~k~, that would not affect anything.  
 \begin{pcodeblock}
    [sp.Rational(1, 2*k) for k in range(1,6)]
 \end{pcodeblock}

 The second approach is to use \pcode~sp.sequence()~.  This is different in
 various ways:
 \begin{itemize}
  \item Here the variable \pcode~k~ needs to be a SymPy symbol with no
   assigned value.  
  \item The range of the variable (which includes \pcode~k = 5~ but
   not \pcode~k = 6~) is now written as `(k, 1, 5)`, matching
   traditional mathematical notation but not the usual Python
   convention for ranges. 
  \item In this example the upper limit for \pcode~k~ is the explicit
   integer $5$, but we could also do something like 
   \pcode~sp.sequence(1/(2 * k), (k, 1, n))~ where \pcode~n~ is a
   SymPy symbol with no assigned value.  This would not be possible
   with an ordinary list comprehension. 
 \end{itemize}

 We can convert a SymPy sequence to an ordinary list using the
 \pcode~list()~ function. 
 \begin{pcodeblock}
    k = sp.symbols('k')
    u = sp.sequence(1/(2 * k), (k, 1, 5))
    display(u)
    display(list(u))
 \end{pcodeblock}
%
 \item\label{sum} 
  For an ordinary Python list, we can use `sum()` to find the sum of the
  terms.
  \begin{pcodeblock}
    u = list(10 ** k for k in range(5))
    display(u)
    display(sum(u))
    display(sum(10 ** k for k in range(5)))   
  \end{pcodeblock}
%
 \item\label{symbolicsum} 
  For a SymPy sequence in which the upper and/or lower limits are
  symbols, we need to use `sp.Sum()` instead of `sum()`.  For example,
  we can get the geometric progression formula for $\sum_{k=0}^n 3^{-k}$
  as follows: 
  \begin{pcodeblock}
   k, n = sp.symbols('k n', integer=True)
   sp.Sum(3**(-k),(k,0,n)).doit()
  \end{pcodeblock}
  Note that the naming and behaviour is a bit inconsistent with the
  naming and behaviour for integrals.  The function \pcode+sp.Sum()+
  always returns an unevaluated sum which is displayed as something
  like $\sum_{k=0}^n3^{-k}$; it is always necessary to use
  \pcode+.doit()+ to find the actual value of the sum.
%
 \item\label{infinitesum} We can also handle some infinite sums.  For
  example, it is a well-known result that
  $\sum_{k=1}^\infty k^{-2}=\pi^2/6$, and we can do this in sympy by
  entering \pcode+sp.Sum(k**(-2),(k,1,sp.oo)).doit()+.
\end{secenum}

\section{Some common problems}\label{sec-problems}

\begin{secenum}
 \item\label{symbolsagain} All abstract symbols need to be introduced using somthing
    like \pcode+x = sp.symbols('x')+.  If you try to use $x$ without
    having declared it, you will get an error like
    \pcode+name 'x' is not defined+.  It can also happen that you give
    $x$ a value, perhaps by entering $x = 42$, and later you forget
    that you did that, and try to treat $x$ as an abstract symbol.  
    This might give you any of a range of different error messages, or
    it might just give you an answer that is incorrect or meaningless
    but with no error message.  You can check the status of $x$ by
    entering \pcode+type(x)+.  If $x$ has been declared as an abstract
    symbol, then \pcode+type(x)+ should give
    \pcode+sympy.core.symbol.Symbol+.  If $x$ has been set equal to
    $42$, then \pcode+type(x)+ should give \pcode+int+.  If $x$ has
    not been introduced in any way, then \pcode+type(x)+ will give an
    error.  If $x$ has been given a value that is no longer relevant,
    you can enter \pcode+del x+ to remove it.
%
 \item\label{rememberstar} Remember \pcode~*~'s for multiplication.  
  \begin{itemize}
   \item Enter $2(x+y)$ as \pcode~2*(x+y)~.  If you instead enter 
    \pcode~2(x+y)~ then you may get a cryptic error message saying 
    \pcode+'int' object is not callable+.
   \item Enter $xy$ as $x*y$.  If you instead enter \pcode+xy+ then
    you will usually get an error message
    \pcode+name 'xy' is not defined+.
   \item For the product of $a$ and $x+y$, enter \pcode~a*(x+y)~.
    To evaluate a function $f(t)$ at the point $t=x+y$, enter
    $f(x+y)$.  If $a$ is an object for which \pcode~a*(x+y)~ makes
    sense, then $a(x+y)$ will not make sense and will give an error.
    If $f$ is an object for which $f(x+y)$ makes sense, then $f*(x+y)$
    will not make sense and will give an error.
  \end{itemize}
%
 \item\label{fracbrac} The fraction $\frac{a+b}{c+d}$ must be entered as
  \pcode~(a+b)/(c+d)~ (not \pcode~a+b/c+d~), and
   $\frac{a}{(u+v)(x+y)}$ must be entered as
   \pcode~a/((u+v)*(x+y))~ (not \pcode~a/(u+v)*(x+y)~). 
%
 \item\label{expbrac} $a^{1/4}$ must be entered as
  \pcode~a ** sp.Rational(1,4)~, not \pcode+a ** (1/4)+ or
  \pcode~a ** 1/4~. 
%
 \item\label{shape} Always use round brackets (not square or curly
         ones) for algebraic grouping.  Square brackets are
         used for lists or for subscripts (e.g. $x_5$ is
         \pcode~x[5]~).  Curly brackets are used for sets or Python
         dictionaries. Using the wrong kind of brackets will usually
         generate an error. 
%
 \item\label{removedefs} Remember to remove definitions when you
  have finished with  them, by using \pcode+del+ or clicking the
  \pcode~Restart~ button (but remember that you need to re-execute 
  your \pcode+import+ cells after restarting).  If you enter an expression like $(x+y)^5$
  and SymPy converts it to something completely unrelated like
  $(\sin(\theta)+e^{-u})^5$, the most likely reason is that
  you previously defined \pcode~x=sp.sin(theta);~ and
  \pcode~y=sp.exp(-u);~, and you forgot to remove these
  definitions.  You can enter 
  \pcode+x,y = sp.symbols('x y')+ to reset $x$ and $y$ to be 
  symbols with no assigned value.
%
 \item\label{badexp} Remember the constant $e=2.71828\dots$ must be
  entered as \pcode+sp.E+ or \pcode~sp.exp(1)~, and $e^x$ must be entered as
  \pcode~exp(x)~.  If you enter \pcode+e ** x+ instead of \pcode+sp.exp(x)+,
  you will most often get an error message
  \pcode+name 'e' is not defined+.  However, if you happen to have
  defined $e$, you might get some other behaviour.
%
 \item\label{badpiagain} Remember that the constant $\pi=3.14159\dots$
  must be entered as \pcode~sp.pi~, not \pcode~pi~.  If you try to use
  \pcode~pi~ instead of \pcode~sp.pi~ then you will usually get an
  error \pcode~name 'pi' is not defined~.  If you want to be able to
  use \pcode~pi~ with no prefix, you can enter
  \pcode~from sympy import pi~ at the top of your notebook.
\end{secenum}

\end{document}
